---
title: "Eelgrass Carbon Stock Analysis"
subtitle: "Sediment Carbon Stocks in Newfoundland Eelgrass Meadows"
date: today
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(readr)
library(knitr)
library(sf)
library(gstat)
library(sp)
library(tidyr)
library(BlueCarbon)

source("00_config.R")

theme_set(theme_bw(base_size = 12))
core_colour <- "#2E8B57"
comp_colour <- "#e76f51"
```

## What this report covers

This report walks through a complete analysis of sediment carbon stocks in an
eelgrass (*Zostera marina*) meadow. The workflow has four main stages:

1. **Raw data** — Load field measurements of soil organic carbon (SOC) and bulk
   density (BD) from sediment cores, and compute raw carbon stocks.
2. **Compaction correction** — Percussion coring compresses the sediment. We
   measure the compression in the field and use the `BlueCarbon` R package to
   stretch cores back to their true depths.
3. **Depth harmonization** — Cores are sampled at different depth intervals.
   We use spline interpolation to estimate SOC and BD at a common set of
   standard depths so cores can be compared directly.
4. **Spatial analysis** — Kriging interpolates carbon stocks across the meadow
   using spatial autocorrelation, with and without water depth as a predictor.

**How carbon stock is calculated:**
Carbon stock (kg C/m²) = SOC (g/kg) × bulk density (g/cm³) × layer thickness (cm) ÷ 100.
In words: higher organic carbon concentration, denser sediment, and thicker
layers all mean more carbon stored per square metre.

## Project overview

```{r}
locations <- read_csv(LOCATIONS_FILE, show_col_types = FALSE)
samples   <- read_csv(SAMPLES_FILE,   show_col_types = FALSE)

tibble(
  Metric = c("Site", "Number of cores", "Total samples",
             "Depth intervals", "Maximum core depth (cm)"),
  Value  = c(SITE_NAME,
             as.character(nrow(locations)),
             as.character(nrow(samples)),
             as.character(nrow(DEPTH_INTERVALS)),
             as.character(max(samples$depth_bottom_cm)))
) |> kable(col.names = c("", ""), align = c("l", "r"))
```

# Raw data {#sec-raw}

```{r data-prep}
`%||%` <- function(a, b) if (!is.null(a)) a else b

cores_raw <- samples |>
  left_join(locations, by = "core_id") |>
  mutate(
    depth_cm           = (depth_top_cm + depth_bottom_cm) / 2,
    layer_thickness_cm = depth_bottom_cm - depth_top_cm,
    bd_estimated       = is.na(bulk_density_g_cm3),
    bulk_density_g_cm3 = if_else(
      bd_estimated,
      sapply(stratum, function(s) BD_DEFAULTS[[s]] %||% NA_real_),
      bulk_density_g_cm3
    ),
    carbon_stock_kg_m2 = (soc_g_kg * bulk_density_g_cm3 * layer_thickness_cm) / 100
  )
```

## Core locations

```{r fig.cap="Core locations coloured by water depth. Deeper sites have longer water columns and reduced light availability.", fig.height=5}
ggplot(
  cores_raw |> distinct(core_id, longitude, latitude, water_depth_m),
  aes(x = longitude, y = latitude, colour = water_depth_m)
) +
  geom_point(size = 5) +
  geom_text(aes(label = core_id), nudge_y = 0.0006, size = 3, colour = "grey30") +
  scale_colour_viridis_c(option = "C", name = "Water\ndepth (m)") +
  coord_sf() +
  labs(title = "Core locations", subtitle = SITE_NAME,
       x = "Longitude", y = "Latitude")
```

## Measured SOC profiles

```{r fig.cap="Soil organic carbon concentration measured at each depth interval. Each line represents one core.", fig.height=5}
ggplot(
  cores_raw |> arrange(core_id, depth_cm),
  aes(x = soc_g_kg, y = depth_cm, group = core_id, colour = core_id)
) +
  geom_path(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_y_reverse(name = "Depth (cm)") +
  scale_x_continuous(name = "SOC (g/kg)") +
  labs(title = "Raw SOC depth profiles", colour = "Core")
```

## Measured bulk density profiles

```{r fig.cap="Bulk density increases with depth as sediments compact under overburden pressure.", fig.height=5}
ggplot(
  cores_raw |> arrange(core_id, depth_cm),
  aes(x = bulk_density_g_cm3, y = depth_cm, group = core_id, colour = core_id)
) +
  geom_path(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_y_reverse(name = "Depth (cm)") +
  scale_x_continuous(name = expression("Bulk density (g cm"^{-3}*")")) +
  labs(title = "Raw bulk density profiles", colour = "Core")
```

## Measured carbon stocks

```{r fig.cap="Carbon stock per sample layer (SOC x BD x layer thickness / 100). Thicker layers contribute more to total stock.", fig.height=5}
ggplot(
  cores_raw |> arrange(core_id, depth_cm),
  aes(x = carbon_stock_kg_m2, y = depth_cm, group = core_id, colour = core_id)
) +
  geom_path(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_y_reverse(name = "Depth (cm)") +
  scale_x_continuous(name = expression("Carbon stock (kg C m"^{-2}*")")) +
  labs(title = "Raw carbon stock profiles", colour = "Core")
```

```{r fig.cap="Total raw carbon stock per core, summed across all measured layers."}
core_totals_raw <- cores_raw |>
  group_by(core_id) |>
  summarise(total_stock = sum(carbon_stock_kg_m2, na.rm = TRUE), .groups = "drop")

ggplot(core_totals_raw, aes(x = reorder(core_id, -total_stock), y = total_stock)) +
  geom_col(fill = core_colour, alpha = 0.8) +
  labs(title = "Total carbon stock per core (raw)",
       x = NULL,
       y = expression("Carbon stock (kg C m"^{-2}*")"))
```

## Raw data summary

```{r}
cores_raw |>
  group_by(core_id) |>
  summarise(
    `Max depth (cm)`  = max(depth_bottom_cm),
    `n layers`        = n(),
    `SOC range (g/kg)` = paste0(round(min(soc_g_kg), 1), " - ", round(max(soc_g_kg), 1)),
    `BD range (g/cm3)` = paste0(round(min(bulk_density_g_cm3), 2), " - ",
                                round(max(bulk_density_g_cm3), 2)),
    `Total stock (kg/m2)` = round(sum(carbon_stock_kg_m2, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  kable(caption = "Per-core summary of raw measurements.")
```


# Compaction correction {#sec-compaction}

When you push a coring tube into soft sediment, the core gets compressed — the
sediment inside the tube ends up shorter than what was actually in the ground.
This means our measured depths are too shallow, and bulk density appears higher
than it really is (the same mass of sediment is squeezed into less space).

We can measure how much compression happened in the field, then use the
`BlueCarbon` R package to stretch the core back to its true length and adjust
bulk density accordingly.

## Compaction estimation

Each core has three field measurements:

- **Sampler length** — total length of the coring tube
- **Internal distance** — distance from tube top to sediment surface *inside* the tube
- **External distance** — distance from tube top to sediment surface *outside* the tube

The compaction ratio is the recovered core length divided by the true
sediment penetration depth. A ratio of 1.0 means no compaction; lower
values indicate more compression.

```{r compaction-estimate}
comp_meas <- read_csv(COMPACTION_FILE, show_col_types = FALSE)

comp_pct <- comp_meas |>
  mutate(
    recovered_length = sampler_length - internal_distance,
    true_length      = sampler_length - external_distance,
    compaction_ratio = recovered_length / true_length,
    compaction       = 1 - compaction_ratio
  )
```

```{r fig.cap="Compaction ratio per core. A ratio of 1.0 means no compaction. Lower values indicate more compression during coring. Typical values for percussion coring in soft eelgrass sediments range from 0.75-0.95.", fig.height=4}
ggplot(comp_pct, aes(x = reorder(core_id, compaction_ratio), y = compaction_ratio)) +
  geom_col(fill = comp_colour, alpha = 0.85, width = 0.6) +
  geom_text(aes(label = round(compaction_ratio, 3)),
            vjust = -0.5, size = 4) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  labs(title = "Compaction ratio per core",
       subtitle = "Ratio = recovered length / true length. Values < 1 indicate compression.",
       x = NULL, y = "Compaction ratio") +
  ylim(0, 1.1)
```

```{r}
comp_pct |>
  select(
    core_id,
    `True length (cm)` = true_length,
    `Recovered length (cm)` = recovered_length,
    `Compaction ratio` = compaction_ratio
  ) |>
  mutate(`Compaction ratio` = round(`Compaction ratio`, 3)) |>
  kable(caption = "Compaction field measurements and compaction ratio per core.")
```

## Decompaction

The `decompact()` function stretches core depths back to their true
(pre-compaction) positions and adjusts bulk density downward — the same mass
of sediment now occupies a larger volume.

```{r decompact}
samples_bc <- samples |>
  left_join(select(comp_pct, core_id, compaction), by = "core_id") |>
  mutate(compaction = coalesce(compaction, 0), soc_pct = soc_g_kg / 10)

names(samples_bc)[names(samples_bc) == "depth_top_cm"]       <- "mind"
names(samples_bc)[names(samples_bc) == "depth_bottom_cm"]    <- "maxd"
names(samples_bc)[names(samples_bc) == "bulk_density_g_cm3"] <- "dbd"

decompacted <- decompact(
  samples_bc,
  core = "core_id", compaction = "compaction",
  mind = "mind", maxd = "maxd", dbd = "dbd"
)
```

### Depth correction

```{r fig.cap="Each bar shows one sample layer. Raw depths (blue, left) are expanded to corrected depths (orange, right) after decompaction. The most compacted cores show the largest depth shifts at the base.", fig.height=6, fig.width=10}
# Compute corrected depths directly from raw samples and compaction ratios
# corrected_depth = raw_depth / compaction_ratio (core expands to true length)
depth_for_plot <- samples |>
  left_join(select(comp_pct, core_id, compaction_ratio), by = "core_id") |>
  mutate(
    corr_top    = depth_top_cm / compaction_ratio,
    corr_bottom = depth_bottom_cm / compaction_ratio
  )

raw_layers <- data.frame(
  core_id = depth_for_plot$core_id,
  top     = depth_for_plot$depth_top_cm,
  bottom  = depth_for_plot$depth_bottom_cm,
  method  = "Raw (compressed)",
  xmin    = 0, xmax = 0.45,
  compaction_ratio = depth_for_plot$compaction_ratio
)
corr_layers <- data.frame(
  core_id = depth_for_plot$core_id,
  top     = depth_for_plot$corr_top,
  bottom  = depth_for_plot$corr_bottom,
  method  = "Corrected (decompacted)",
  xmin    = 0.55, xmax = 1,
  compaction_ratio = depth_for_plot$compaction_ratio
)
depth_long <- rbind(raw_layers, corr_layers) |>
  mutate(core_label = paste0(core_id, "\n(ratio: ", round(compaction_ratio, 3), ")"))

ggplot(depth_long, aes(xmin = xmin, xmax = xmax, ymin = top, ymax = bottom, fill = method)) +
  geom_rect(alpha = 0.7, colour = "white", linewidth = 0.3) +
  scale_y_reverse(name = "Depth (cm)") +
  scale_fill_manual(values = c("Raw (compressed)" = "#6baed6",
                                "Corrected (decompacted)" = comp_colour)) +
  facet_grid(~ reorder(core_label, compaction_ratio)) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank()) +
  labs(title = "Depth intervals: raw vs decompacted",
       subtitle = "Cores ordered by compaction ratio. Corrected depths are deeper (sediment expanded).",
       x = NULL, fill = NULL)
```

### Bulk density correction

```{r fig.cap="Decompaction reduces bulk density because the same sediment mass is distributed across a larger (true) volume. The correction is proportional to compaction — more compressed cores see larger BD reductions.", fig.height=5}
# Compute corrected BD directly: bd_corrected = bd_raw * compaction_ratio
# (same mass spread over expanded volume)
bd_for_plot <- samples |>
  left_join(select(comp_pct, core_id, compaction_ratio), by = "core_id") |>
  mutate(
    depth_mid_raw  = (depth_top_cm + depth_bottom_cm) / 2,
    depth_mid_corr = ((depth_top_cm / compaction_ratio) + (depth_bottom_cm / compaction_ratio)) / 2,
    bd_corr        = bulk_density_g_cm3 * compaction_ratio
  )

bd_long <- bd_for_plot |>
  pivot_longer(cols = c(bulk_density_g_cm3, bd_corr),
               names_to = "method", values_to = "bd") |>
  mutate(
    depth_mid = if_else(method == "bulk_density_g_cm3", depth_mid_raw, depth_mid_corr),
    method = if_else(method == "bulk_density_g_cm3", "Raw", "Decompacted")
  )

ggplot(bd_long, aes(x = bd, y = depth_mid, colour = method, group = interaction(core_id, method))) +
  geom_path(linewidth = 0.6, alpha = 0.7) +
  geom_point(size = 1.5) +
  scale_y_reverse(name = "Depth (cm)") +
  scale_colour_manual(values = c("Raw" = "#6baed6", "Decompacted" = comp_colour)) +
  facet_wrap(~core_id) +
  labs(title = "Bulk density: raw vs decompacted",
       subtitle = "Decompacted BD is lower — same mass over expanded depth",
       x = expression("Bulk density (g cm"^{-3}*")"),
       colour = NULL) +
  theme(legend.position = "top")
```

```{r compute-decompacted-h}
decompacted_with_h <- estimate_h(
  decompacted,
  core = "core_id",
  mind = "mind_corrected",
  maxd = "maxd_corrected"
)
```


# Depth harmonization {#sec-harmonized}

Cores are sampled at different depth intervals depending on how the core was
sectioned in the lab. To compare across cores and compute site-level averages,
we need all profiles on the same depth scale. Harmonization uses smooth curve
fitting (spline interpolation) to estimate SOC and bulk density at a common
set of standard depth intervals.

The harmonization is applied to the **decompacted** cores — depth intervals
and bulk density have already been corrected for compaction in the previous step.

**Standard depth intervals:**

```{r}
DEPTH_INTERVALS |>
  mutate(Label = paste0(depth_top, "-", depth_bottom, " cm")) |>
  select(Label, `Midpoint (cm)` = depth_midpoint, `Thickness (cm)` = thickness_cm) |>
  kable()
```

**Method:**

- **Within measured range:** smooth curve fitting (monotone spline interpolation)
  that follows the shape of measured profiles without creating artificial peaks
  or dips between data points.
- **Beyond measured range:** if SOC clearly decreases with depth, the curve
  continues that declining trend (exponential decay). Otherwise, the deepest
  measured value is carried downward. Extrapolation is limited to 2.5× the
  maximum measured depth.

```{r harmonize}
fit_hybrid_profile <- function(depths, values, targets) {
  if (length(depths) < 2) return(rep(NA_real_, length(targets)))
  max_d <- max(depths)
  fn <- tryCatch(splinefun(depths, values, method = "monoH.FC"), error = function(e) NULL)
  if (is.null(fn)) return(rep(NA_real_, length(targets)))
  decay <- NULL
  rho <- tryCatch(cor(depths, values, method = "spearman", use = "complete.obs"),
                  error = function(e) NA_real_)
  if (!is.na(rho) && rho < -0.3 && length(depths) >= 3)
    try(decay <- lm(log(values + 0.1) ~ depths), silent = TRUE)
  sapply(targets, function(d) {
    if (d <= max_d) fn(d)
    else if (d > max_d * 2.5) NA_real_
    else if (!is.null(decay) && coef(decay)[2] < 0)
      exp(predict(decay, newdata = data.frame(depths = d))) - 0.1
    else values[which.max(depths)]
  })
}

# Use decompacted cores as input
cores_for_harm <- decompacted_with_h |>
  left_join(select(locations, core_id, stratum, latitude, longitude, water_depth_m),
            by = "core_id") |>
  mutate(
    depth_cm           = (mind_corrected + maxd_corrected) / 2,
    bulk_density_g_cm3 = dbd_corrected
  )

cores_qa <- cores_for_harm |>
  filter(!is.na(depth_cm), !is.na(soc_g_kg), !is.na(bulk_density_g_cm3)) |>
  arrange(core_id, depth_cm)

results <- lapply(unique(cores_qa$core_id), function(id) {
  sub <- cores_qa |> filter(core_id == id) |> arrange(depth_cm)
  if (nrow(sub) < 2) return(NULL)
  tryCatch({
    depths   <- sub$depth_cm
    soc_pred <- fit_hybrid_profile(depths, sub$soc_g_kg, DEPTH_MIDPOINTS)
    bd_fn    <- splinefun(depths, sub$bulk_density_g_cm3, method = "monoH.FC")
    bd_last  <- sub$bulk_density_g_cm3[which.max(depths)]
    bd_pred  <- sapply(DEPTH_MIDPOINTS, function(d) if (d <= max(depths)) bd_fn(d) else bd_last)
    data.frame(
      core_id = id, stratum = sub$stratum[1],
      latitude = sub$latitude[1], longitude = sub$longitude[1],
      water_depth_m = sub$water_depth_m[1],
      depth_cm_midpoint = DEPTH_MIDPOINTS,
      thickness_cm = DEPTH_INTERVALS$thickness_cm,
      soc_harmonized = pmax(0, soc_pred),
      bd_harmonized = pmax(0, bd_pred),
      is_extrapolated = DEPTH_MIDPOINTS > max(depths)
    )
  }, error = function(e) NULL)
})

cores_harmonized <- bind_rows(results) |>
  mutate(carbon_stock_kg_m2 = (soc_harmonized * bd_harmonized * thickness_cm) / 100) |>
  filter(!is.na(carbon_stock_kg_m2))
```

## Harmonized SOC profiles

```{r fig.cap="Harmonized SOC at standard depth midpoints. Open circles indicate extrapolated values beyond the measured depth range.", fig.height=5}
ggplot(cores_harmonized,
  aes(x = soc_harmonized, y = depth_cm_midpoint,
      group = core_id, colour = core_id)) +
  geom_path(linewidth = 0.7) +
  geom_point(aes(shape = is_extrapolated), size = 2.5) +
  scale_y_reverse(name = "Depth midpoint (cm)") +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 1),
                     labels = c("Interpolated", "Extrapolated")) +
  labs(title = "Harmonized SOC profiles (from decompacted cores)",
       x = "SOC (g/kg)", colour = "Core", shape = "Method")
```

## Harmonized bulk density profiles

```{r fig.cap="Harmonized bulk density at standard depth midpoints. Values reflect decompacted (corrected) bulk densities.", fig.height=5}
ggplot(cores_harmonized,
  aes(x = bd_harmonized, y = depth_cm_midpoint,
      group = core_id, colour = core_id)) +
  geom_path(linewidth = 0.7) +
  geom_point(aes(shape = is_extrapolated), size = 2.5) +
  scale_y_reverse(name = "Depth midpoint (cm)") +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 1),
                     labels = c("Interpolated", "Extrapolated")) +
  labs(title = "Harmonized bulk density profiles (decompacted)",
       x = expression("Bulk density (g cm"^{-3}*")"),
       colour = "Core", shape = "Method")
```

## Processing pipeline comparison

This section shows how SOC values and total carbon stocks change through
each stage of the workflow — from raw measurements, through compaction
correction, to harmonized standard depths.

```{r fig.cap="SOC comparison across all three processing stages: raw measurements (blue points), harmonized profile (coloured line + diamonds at standard depths).", fig.height=6, fig.width=10}
ggplot() +
  # Raw measurements (original compressed depths)
  geom_point(data = cores_raw,
             aes(x = soc_g_kg, y = depth_cm),
             colour = "#6baed6", size = 2, alpha = 0.5) +
  # Harmonized profile — connect the diamonds
  geom_path(data = cores_harmonized |> arrange(core_id, depth_cm_midpoint),
            aes(x = soc_harmonized, y = depth_cm_midpoint, colour = core_id),
            linewidth = 0.7) +
  # Harmonized values at standard depths
  geom_point(data = cores_harmonized,
             aes(x = soc_harmonized, y = depth_cm_midpoint, colour = core_id),
             shape = 18, size = 3.5) +
  geom_hline(yintercept = DEPTH_INTERVALS$depth_top, linetype = "dashed",
             alpha = 0.3) +
  scale_y_reverse(name = "Depth (cm)") +
  facet_wrap(~core_id, scales = "free_x") +
  labs(title = "Raw vs decompacted vs harmonized SOC",
       subtitle = "Blue points = raw, lines = decompacted spline, diamonds = harmonized at standard depths",
       x = "SOC (g/kg)", colour = "Core") +
  theme(legend.position = "none")
```

```{r fig.cap="Total carbon stock per core at each processing stage. Raw = as measured. Decompacted = after correcting for coring compression (depths expanded, BD reduced). Harmonized = decompacted values interpolated to standard depth intervals.", fig.height=5}
# Decompacted totals (intermediate stage)
cores_decomp_stock <- decompacted_with_h |>
  mutate(
    layer_thickness = maxd_corrected - mind_corrected,
    stock = soc_g_kg * dbd_corrected * layer_thickness / 100
  )
core_totals_decomp <- cores_decomp_stock |>
  group_by(core_id) |>
  summarise(decompacted = sum(stock, na.rm = TRUE), .groups = "drop")

# Harmonized totals
core_totals_harm <- cores_harmonized |>
  group_by(core_id) |>
  summarise(harmonized = sum(carbon_stock_kg_m2, na.rm = TRUE), .groups = "drop")

three_way <- core_totals_raw |>
  rename(raw = total_stock) |>
  left_join(core_totals_decomp, by = "core_id") |>
  left_join(core_totals_harm, by = "core_id") |>
  pivot_longer(cols = c(raw, decompacted, harmonized),
               names_to = "stage", values_to = "stock") |>
  mutate(stage = factor(stage, levels = c("raw", "decompacted", "harmonized")))

ggplot(three_way, aes(x = core_id, y = stock, fill = stage)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(
    values = c("raw" = "grey60", "decompacted" = comp_colour, "harmonized" = core_colour),
    labels = c("Raw (as measured)", "Decompacted (depth + BD corrected)",
               "Harmonized (standard depths)")
  ) +
  labs(title = "Total carbon stock across processing stages",
       subtitle = "Each bar shows the total 0-100 cm carbon stock at a different stage of the workflow",
       x = NULL,
       y = expression("Carbon stock (kg C m"^{-2}*")"),
       fill = NULL) +
  theme(legend.position = "top")
```


# Carbon stock estimates {#sec-estimates}

## Per-depth summary

```{r}
depth_summary <- cores_harmonized |>
  group_by(`Depth interval` = paste0(
    DEPTH_INTERVALS$depth_top[match(depth_cm_midpoint, DEPTH_MIDPOINTS)],
    "-",
    DEPTH_INTERVALS$depth_bottom[match(depth_cm_midpoint, DEPTH_MIDPOINTS)],
    " cm")) |>
  summarise(
    `Midpoint (cm)` = first(depth_cm_midpoint),
    `n cores`        = n_distinct(core_id),
    `Mean SOC (g/kg)` = round(mean(soc_harmonized, na.rm = TRUE), 1),
    `Mean BD (g/cm3)` = round(mean(bd_harmonized, na.rm = TRUE), 2),
    `Mean stock (kg/m2)` = round(mean(carbon_stock_kg_m2, na.rm = TRUE), 2),
    `SD (kg/m2)`     = round(sd(carbon_stock_kg_m2, na.rm = TRUE), 2),
    `SE (kg/m2)`     = round(sd(carbon_stock_kg_m2, na.rm = TRUE) / sqrt(n_distinct(core_id)), 2),
    `% extrapolated`  = round(mean(is_extrapolated) * 100, 0),
    .groups = "drop"
  ) |>
  arrange(`Midpoint (cm)`)

depth_summary |>
  select(-`Midpoint (cm)`) |>
  kable(caption = "Harmonized carbon stocks by depth interval (from decompacted cores).")
```

## Total carbon stock (0-100 cm)

```{r}
depth_stats <- cores_harmonized |>
  group_by(depth_cm_midpoint) |>
  summarise(
    mean_stock = mean(carbon_stock_kg_m2, na.rm = TRUE),
    se_stock   = sd(carbon_stock_kg_m2, na.rm = TRUE) / sqrt(n_distinct(core_id)),
    .groups = "drop"
  )

total_mean  <- sum(depth_stats$mean_stock)
total_se    <- sqrt(sum(depth_stats$se_stock^2))
ci90_lower  <- total_mean - 1.645 * total_se
ci90_upper  <- total_mean + 1.645 * total_se

tibble(
  Metric = c("Mean total stock", "Standard error",
             "90% CI lower bound", "90% CI upper bound",
             "Equivalent (Mg C/ha)"),
  Value  = c(
    paste0(round(total_mean, 2), " kg C/m2"),
    paste0(round(total_se, 2), " kg C/m2"),
    paste0(round(ci90_lower, 2), " kg C/m2"),
    paste0(round(ci90_upper, 2), " kg C/m2"),
    paste0(round(total_mean * 10, 1), " Mg C/ha")
  )
) |> kable(col.names = c("", ""),
           caption = "Site-level carbon stock estimate (0-100 cm, from decompacted + harmonized cores).")
```

```{r fig.cap="Mean carbon stock by depth interval with 90% confidence intervals. Estimates are based on decompacted, harmonized cores.", fig.height=4.5}
plot_data <- cores_harmonized |>
  group_by(depth_cm_midpoint) |>
  summarise(
    mean_stock = mean(carbon_stock_kg_m2, na.rm = TRUE),
    se_stock   = sd(carbon_stock_kg_m2, na.rm = TRUE) / sqrt(n_distinct(core_id)),
    ci90_lower = mean_stock - qt(0.95, n_distinct(core_id) - 1) * se_stock,
    ci90_upper = mean_stock + qt(0.95, n_distinct(core_id) - 1) * se_stock,
    .groups = "drop"
  )

ggplot(plot_data, aes(x = mean_stock,
                      y = factor(depth_cm_midpoint))) +
  geom_col(fill = core_colour, alpha = 0.7) +
  geom_errorbarh(aes(xmin = pmax(0, ci90_lower), xmax = ci90_upper), height = 0.3) +
  scale_y_discrete(name = "Depth midpoint (cm)",
                   limits = rev(as.character(DEPTH_MIDPOINTS))) +
  labs(title = "Mean carbon stock by depth (90% CI)",
       subtitle = SITE_NAME,
       x = expression("Carbon stock (kg C m"^{-2}*")"))
```

## Per-core breakdown

```{r fig.cap="Stacked carbon stock by depth interval for each core. Colour intensity increases with depth.", fig.height=5}
ggplot(cores_harmonized,
  aes(x = core_id, y = carbon_stock_kg_m2,
      fill = factor(depth_cm_midpoint))) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(
    values = c("7.5" = "#a1d99b", "22.5" = "#74c476",
               "40" = "#31a354", "75" = "#006d2c"),
    name = "Depth\nmidpoint (cm)") +
  labs(title = "Carbon stock breakdown by core and depth",
       x = NULL,
       y = expression("Carbon stock (kg C m"^{-2}*")")) +
  theme(legend.position = "right")
```


# Spatial analysis — Kriging {#sec-kriging}

With only `r n_distinct(cores_harmonized$core_id)` cores, we can still
estimate spatial patterns by assuming that locations close together tend to
have similar carbon stocks. Kriging formalises this idea: it fits a model of
how similarity decays with distance (the **variogram**), then uses that model
to predict stocks at unsampled locations and estimate the prediction
uncertainty.

Two approaches are shown:

1. **Ordinary kriging** — predicts based on distance to nearby cores only.
2. **Universal kriging** — also accounts for water depth, so areas at similar
   depths to high-carbon cores receive higher predicted stocks.

```{r kriging-setup}
core_stocks <- cores_harmonized |>
  group_by(core_id, latitude, longitude, water_depth_m) |>
  summarise(total_stock_kg_m2 = sum(carbon_stock_kg_m2, na.rm = TRUE), .groups = "drop")

cores_sf  <- st_as_sf(core_stocks, coords = c("longitude", "latitude"), crs = 4326)
cores_utm <- st_transform(cores_sf, crs = 32621)
coords_utm <- st_coordinates(cores_utm)
cores_utm$x <- coords_utm[, 1]
cores_utm$y <- coords_utm[, 2]

bbox   <- st_bbox(cores_utm)
buffer <- 200
grid_pts <- expand.grid(
  x = seq(bbox["xmin"] - buffer, bbox["xmax"] + buffer, by = 25),
  y = seq(bbox["ymin"] - buffer, bbox["ymax"] + buffer, by = 25)
)
grid_sf <- st_as_sf(grid_pts, coords = c("x", "y"), crs = 32621)

sp_cores <- as(cores_utm, "Spatial")
sp_grid  <- as(grid_sf, "Spatial")

idw_depth <- idw(water_depth_m ~ 1, locations = sp_cores, newdata = sp_grid, idp = 2)
grid_pts$water_depth_m <- idw_depth$var1.pred

sp_cores_df <- SpatialPointsDataFrame(
  coords = cbind(cores_utm$x, cores_utm$y),
  data = data.frame(total_stock = cores_utm$total_stock_kg_m2,
                    water_depth_m = cores_utm$water_depth_m),
  proj4string = CRS("+init=epsg:32621")
)

core_plot_pts <- data.frame(x = cores_utm$x, y = cores_utm$y,
                            core_id = core_stocks$core_id)
```

## Variogram analysis

```{r fig.cap="Variogram: each point shows how different pairs of cores are at a given separation distance. The fitted line is the model kriging uses to predict at new locations. The flat part (sill) is the overall variance; the distance where the curve levels off (range) is how far apart cores need to be before they stop telling us about each other.", fig.height=4.5}
vg_ok <- variogram(total_stock ~ 1, data = sp_cores_df)

vg_fit_ok <- tryCatch(
  fit.variogram(vg_ok, vgm(psill = var(core_stocks$total_stock_kg_m2),
                            model = "Sph",
                            range = max(dist(coords_utm)) / 2,
                            nugget = 0.1)),
  error = function(e) {
    fit.variogram(vg_ok, vgm(psill = var(core_stocks$total_stock_kg_m2),
                              model = "Exp",
                              range = max(dist(coords_utm)) / 3,
                              nugget = 0.1))
  }
)

plot(vg_ok, vg_fit_ok, main = "Ordinary Kriging — Variogram")
```

```{r}
tibble(
  Component = vg_fit_ok$model,
  Sill      = round(vg_fit_ok$psill, 2),
  Range     = round(vg_fit_ok$range, 0)
) |> kable(caption = "Fitted variogram parameters.")
```

## Ordinary kriging

```{r ok-predict}
sp_grid_ok <- SpatialPointsDataFrame(
  coords = cbind(grid_pts$x, grid_pts$y),
  data = grid_pts,
  proj4string = CRS("+init=epsg:32621")
)

ok_result <- krige(total_stock ~ 1, locations = sp_cores_df,
                   newdata = sp_grid_ok, model = vg_fit_ok)
grid_pts$ok_pred <- ok_result$var1.pred
grid_pts$ok_var  <- ok_result$var1.var
```

```{r fig.cap="Ordinary kriging prediction of total carbon stock (0-100 cm). Red points show core locations.", fig.height=5.5}
ggplot(grid_pts, aes(x = x, y = y, fill = ok_pred)) +
  geom_tile() +
  scale_fill_viridis_c(option = "D", name = expression("kg C m"^{-2})) +
  geom_point(data = core_plot_pts, colour = "red", size = 3,
             shape = 21, fill = "white", stroke = 0.8) +
  geom_text(data = core_plot_pts, aes(x = x, y = y, label = core_id),
            nudge_y = 40, size = 2.5, colour = "white", inherit.aes = FALSE) +
  coord_equal() +
  labs(title = "Ordinary Kriging — Total Carbon Stock",
       subtitle = SITE_NAME, x = "Easting (m)", y = "Northing (m)")
```

```{r fig.cap="Prediction standard error — uncertainty is lowest near core locations and increases with distance. This map helps identify where additional sampling would most reduce uncertainty.", fig.height=5.5}
ggplot(grid_pts, aes(x = x, y = y, fill = sqrt(ok_var))) +
  geom_tile() +
  scale_fill_viridis_c(option = "A", name = expression("SE (kg C m"^{-2}*")")) +
  geom_point(data = core_plot_pts, colour = "red", size = 3,
             shape = 21, fill = "white", stroke = 0.8) +
  coord_equal() +
  labs(title = "Ordinary Kriging — Prediction Standard Error",
       x = "Easting (m)", y = "Northing (m)")
```

## Universal kriging (water depth as drift)

Universal kriging first accounts for the effect of water depth on carbon stocks
(the trend), then models the remaining spatial pattern in the residuals. This
can improve predictions when carbon stocks vary systematically with depth.

```{r uk-predict}
vg_uk <- variogram(total_stock ~ water_depth_m, data = sp_cores_df)

vg_fit_uk <- tryCatch(
  fit.variogram(vg_uk, vgm(psill = var(core_stocks$total_stock_kg_m2) * 0.5,
                            model = "Sph",
                            range = max(dist(coords_utm)) / 2,
                            nugget = 0.1)),
  error = function(e) {
    fit.variogram(vg_uk, vgm(psill = var(core_stocks$total_stock_kg_m2) * 0.5,
                              model = "Exp",
                              range = max(dist(coords_utm)) / 3,
                              nugget = 0.1))
  }
)

sp_grid_uk <- SpatialPointsDataFrame(
  coords = cbind(grid_pts$x, grid_pts$y),
  data = grid_pts,
  proj4string = CRS("+init=epsg:32621")
)

uk_result <- krige(total_stock ~ water_depth_m, locations = sp_cores_df,
                   newdata = sp_grid_uk, model = vg_fit_uk)
grid_pts$uk_pred <- uk_result$var1.pred
grid_pts$uk_var  <- uk_result$var1.var
```

```{r fig.cap="Universal kriging uses water depth to improve predictions — areas with similar water depth to high-carbon cores receive higher predicted stocks.", fig.height=5.5}
ggplot(grid_pts, aes(x = x, y = y, fill = uk_pred)) +
  geom_tile() +
  scale_fill_viridis_c(option = "D", name = expression("kg C m"^{-2})) +
  geom_point(data = core_plot_pts, colour = "red", size = 3,
             shape = 21, fill = "white", stroke = 0.8) +
  coord_equal() +
  labs(title = "Universal Kriging — Total Carbon Stock (drift: water depth)",
       subtitle = SITE_NAME, x = "Easting (m)", y = "Northing (m)")
```

## Comparison: ordinary vs universal kriging

```{r fig.cap="Difference between universal and ordinary kriging predictions. Positive values (green) indicate areas where accounting for water depth increases the predicted stock.", fig.height=5.5}
grid_pts$diff <- grid_pts$uk_pred - grid_pts$ok_pred

ggplot(grid_pts, aes(x = x, y = y, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9850",
                       midpoint = 0,
                       name = expression(Delta~"kg C m"^{-2})) +
  geom_point(data = core_plot_pts, colour = "black", size = 3,
             shape = 21, fill = "white", stroke = 0.8) +
  coord_equal() +
  labs(title = "Difference: Universal - Ordinary Kriging",
       x = "Easting (m)", y = "Northing (m)")
```

```{r}
tibble(
  Method = c("Ordinary kriging", "Universal kriging"),
  `Mean prediction (kg/m2)` = round(c(mean(grid_pts$ok_pred, na.rm = TRUE),
                                       mean(grid_pts$uk_pred, na.rm = TRUE)), 2),
  `Mean SE (kg/m2)` = round(c(mean(sqrt(grid_pts$ok_var), na.rm = TRUE),
                               mean(sqrt(grid_pts$uk_var), na.rm = TRUE)), 2),
  `Equivalent (Mg C/ha)` = round(c(mean(grid_pts$ok_pred, na.rm = TRUE) * 10,
                                    mean(grid_pts$uk_pred, na.rm = TRUE) * 10), 1)
) |> kable(caption = "Comparison of kriging approaches.")
```

## Caveats

::: {.callout-note}
With only `r n_distinct(core_stocks$core_id)` cores, variogram fitting is
unreliable and kriging predictions carry high uncertainty. These results are
illustrative — treat the spatial patterns as hypotheses to test with additional
sampling, not as definitive carbon stock maps.

Water depth at prediction grid points was approximated by inverse distance
weighting from core locations. For operational use, replace this with a
bathymetry raster.
:::

---

*Rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M")`*
